---
title: "Basic Use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

```{r setup}
library(simplefreqs)
```

## Printing

This package provides a easy to produce frequency tables.  It is built to
print nicely both in the console and in knitted documents (i.e. in Rmarkdown or Quarto).
It should produce attractive console, html, pdf, and word output.

### Console

```{r console}
freq(iris, Species, markdown = F, plot = F)
```

### HTML

```{r html}
freq(iris, Species, markdown = T)
```

### PDF

```{r pdf}
freq(iris, Species, markdown = T)
```

### Word

```{r word}
freq(iris, Species, markdown = T)
```

## Tidyverse

It is designed to be fit into a piped/Tidyverse workflow. 

```{r tidyverse, message=F, error=F, warning=F}
library(dplyr)
```

Thus, the first parameter should be a data.frame (or tibble) and the second 
parameter an unquoted variable from the specified data.frame

```{r basic}
data(storms)
freq(storms, status)
```

or 

```{r basic2}
storms |>
  freq(status)
```

## Other Accepted Inputs

Alternatively, `freq` will detect if you pass a vector as the first parameter and
produce a frequency table from it.

```{r vector}
freq(storms$category)
```

or

```{r vector2}
freq(storms[["category"]])
```

## Assigning Results

Results need not be printed. Results, instead, can be assigned to a 
variable.  The results will be a data.fram with a class of both "simplefreqs_freq" 
and "data.frame".

```{r assigning}
df <- freq(mtcars, cyl)
class(df)
```

This object will continue to print as a `simplefreqs_freqs`.

```{r assigning2}
df <- freq(mtcars, cyl)
df
```

If needed, you can remove the "simplefreqs_freq" class and the object will print 
as a regular data.frame.

```{r assigning3}
df <- freq(mtcars, cyl)
df <- as.data.frame(df)
class(df)
df
```

## Plots
By default, a simple bar chart is produced alongside the frequency table.

```{r plots1}
freq(iris, Species)
```

Plots can be supressed.

```{r plots2}
freq(iris, Species, plot = FALSE)
```

## Sorting

By default, results are sorted in descending order by frequency.

```{r sort1}
freq(iris, Species, sort = TRUE, plot = FALSE)
```

Alternatively, sorting can be supressed. In that case, results will be sorted by 
the variables labels. For character variables, it will be alphabetic, for factors
it will in order of the levels.

```{r sort2}
freq(iris, Species, sort = FALSE, plot = FALSE)
```

# NAs

NAs are, by default, included and treated as any other level of the variable.

```{r nas1}
df <- iris |>
  mutate(Species=ifelse(row_number()<10,NA,Species)) |>
    freq(Species, na.rm = FALSE)
df
```

Alternatively, NAs can be removed. In that case, NAs will not be included in the 
frequency table, but two relevent attribute will be set on the results. "na" will include the number
of NAs removed and na_removed will be set to TRUE.

```{r nas2}

df <- iris |>
  mutate(Species=ifelse(row_number()<10,NA,Species)) |>
    freq(Species, na.rm = TRUE)
df
attr(df, "na", exact = TRUE)
attr(df, "na_removed", exact = TRUE)
```

## Markdown

`simplefreqs::freq` tries to intuit if you are printing to the console 
or are knitting a document. In some cases, you may need to directly specify the 
type of output you want to display using the `markdown` parameter.  Set to `FALSE`
when you wish to print for the console.

```{r mkdown1}
freq(iris, Species, markdown = FALSE)
```

Set to `markdown = TRUE` when you wish to include in a knitted document.

```{r mkdown2}
freq(iris, Species, markdown = TRUE)
```

# Weighting
Results can be optionally weighted by providing a vector of weight values.

```{r wt1}
freq(iris, Species, wt = Sepal.Length)

```

## Optional Formatting
The printed results are highly customizable, both for console output and for HTML
output (PDF and Word output is less customizable).  There are many available options
that control how these tables print.  You can set these options per session (using 
`options(simplefreqs.OPTION = VALUE)`) or by adding the similar lines to your 
project or global .RProfile file.

The exposed customization options are:

- simplefreqs.inner_table_padding
- simplefreqs.table_symbol
- simplefreqs.row_divider_symbol
- simplefreqs.print_table_symbol
- simplefreqs.print_table_total_row
- simplefreqs.print_table_metadata
- simplefreqs.print_header_divider
- simplefreqs.big_mark
- simplefreqs.decimal_digits





